---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# billionaiRe

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/caldwellst/billionaiRe.svg?branch=master)](https://travis-ci.com/caldwellst/billionaiRe)
[![R build status](https://github.com/caldwellst/billionaiRe/workflows/R-CMD-check/badge.svg)](https://github.com/caldwellst/billionaiRe/actions)
<!-- badges: end -->

The goal of billionaiRe is to provide an easy interface for using long format
data to calculate the World Health Organization's Triple Billions.

## Installation

You can install billionaiRe from [GitHub](https://github.com/) with:

```  r
remotes::install_github("caldwellst/billionaiRe", build_vignettes = TRUE)
```

You will need to have already installed the wppdistro package, which is stored in
a private repo and only made public upon request. Please contact caldwellst@who.int
to request access.

# Calculations

The package is built around a set of functions that separately calculate the Billions
for the three Billions separately:

* Healthier Populations (HPOP)
* Health Emergencies Protection (HEP)
* Universal Health Coverage (UHC)

## HPOP Billion calculation

```{r child='vignettes/hpop.rmd'}
```

## UHC Billion calculation

```{r child='vignettes/uhc.rmd'}
```

## HEP Billion calculation

```{r child='vignettes/hep.rmd'}
```
