---
title: "Acceleration scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acceleration scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Acceleration scenarios are version of the future that are ambitious yet achievable by countries. They were created in conjonction with WHO technical units.

Many of the acceleration scenarios calculate multiple basic scenarios, and then compare them with the `scenario_best_of()` function to find the best results.

This table shows a summary of the acceleration scenario by indicator:

```{r show-acceleration-ind, echo=FALSE}
library(billionaiRe)

indicator_df %>% 
  dplyr::mutate(Billion = dplyr::case_when(
    ind == "espar" ~ "UHC and HEP",
    TRUE ~ stringr::str_to_upper(get_ind_billion(ind))
  )) %>% 
  dplyr::filter(!calculated, !is.na(acceleration_scenario)) %>%
  dplyr::select(Billion,
                `Indicator Code` = ind,
                `Indicator Name` = short_name,
                `Scenario` = acceleration_scenario) %>% 
  dplyr::arrange(Billion, `Indicator Code`) %>% 
  knitr::kable("html")
```
