---
title: "HPOP Billion calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hpop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To calculate the HPOP Billion, there are a series of functions made available through
the billionaiRe package:

* `transform_hpop_data()` to transform raw values into normalized values used within the calculations.
* `add_hpop_populations()` to get relevant population groups for each country and indicator.
* `calculate_hpop_contributions()` to calculate indicator level changes and contributions to the Billion.
* `calculate_hpop_billion()` to calculate country-level Billion, adjusting for double counting.

Run in sequence, these can calculate the entire HPOP Billion, or they can be run separately
to produce different outputs as required. Details on the inputs of each function are
available in their individual documentation, but below you can see the quick
and easy Billions calculation done using the sample fake HPOP data provided
in the package, `hpop_df`.

```{r setup_hpop}
library(billionaiRe)

hpop_df %>%
  transform_hpop_data() %>%
  add_hpop_populations() %>%
  calculate_hpop_contributions() %>%
  calculate_hpop_billion()
```

