---
title: "SDG scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SDG scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Sustainable Development Goals (SDG) scenarios are setting the indicators of the Triple Billion on a trajectory to reach the targets set in the [Global indicator framework for the Sustainable Development Goals and targets of the 2030 Agenda for Sustainable Development](https://unstats.un.org/sdgs/indicators/Global%20Indicator%20Framework%20after%202022%20refinement_Eng.pdf).

This results in setting ambitious trajectories for each indicator that is linked to an SDG target or to a relevant framework

```{r show-sdg-ind, echo=FALSE}
library(billionaiRe)

indicator_df %>% 
  dplyr::mutate(Billion = dplyr::case_when(
    ind == "espar" ~ "UHC and HEP",
    TRUE ~ stringr::str_to_upper(get_ind_billion(ind)))) %>% 
  dplyr::filter(!calculated, !is.na(sdg),
                stringr::str_detect(sdg, "^SDG|^WHA")) %>%
  dplyr::select(Billion,
                `Indicator Code` = ind,
                `Indicator Name` = short_name,
                `SDG/WHA code` = sdg,
                `Scenario` = sdg_scenario) %>% 
  dplyr::arrange(Billion, `Indicator Code`) %>% 
  knitr::kable()
```
