---
title: "UHC Billion calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{UHC Billion calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


To calculate the UHC Billion, there are a series of functions made available through
the billionaiRe package:

* `transform_uhc_data()` to transform raw values into normalized values used within the calculations.
* `calculate_uhc_billion()` to calculate average service coverage, financial hardship, and the UHC single measure for each country and year in the data frame..
* `calculate_uhc_contribution()` to calculate country-level Billion for specified beginning and end year.

Run in sequence, these can calculate the entire UHC Billion, or they can be run separately
to produce different outputs as required. Details on the inputs of each function are
available in their individual documentation, but below you can see the quick
and easy Billions calculation done using the the sample fake UHC data provided
in the package, `uhc_df`.

```{r setup_uhc, eval=FALSE}
library(billionaiRe)

uhc_df %>%
  transform_uhc_data() %>%
  calculate_uhc_billion() %>%
  calculate_uhc_contribution(end_year = 2023, pop_year = 2023) %>% 
  dplyr::filter(ind %in% c("uhc_sm", "asc", "fh"),
                year == 2023)
```
